### 2021-04-29 doc/VerticalArrangements.tex
added some new figures. added additional text to explain figures. still a work in progress.

Table \ref{tab:simsum} provides summary statistics across our various simulations.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to calibrate and simulate the effects of mergers in a range of competitive scenarios, including the ones described here.}  The median average wholesale pre-merger price is \$4.6, and the median average retail pre-merger price is \$14.  Because the market size is set to 1, these average prices are equal to total pre-merger expenditures.  Pre-merger HHIs range between 2,814 at the $25^{th}$ percentile to 5,030 at the $75^{th}$, with a median of 3,703. HHIs for horizontal downstream mergers increase by 2,015 points at the median, resulting in a median post-merger HHI equal to 4,914. HHIs for upstream mergers increase by 2,063 points at the median, resulting in a median post-merger HHI equal to 5,238. HHIs for vertical mergers increase by 1,055 points at the median, resulting in a median post-merger HHI equal to 5,995.\footnote{We compute the post-merger HHI for vertical mergers by calculating the merged firms' market share as the sum of all the shares of downstream products that either incorporate the upstream partner's input or are sold by the downstream partner.  }  HHIs for integrated mergers increase by 2,770 points at the median, resulting in a median post-merger HHI equal to 7,280. Many of these markets fall into the span designated by the DOJ/FTC Horizontal Merger Guidelines as ``Highly Concentrated Markets,'' with post-merger HHIs over 2,500 points and HHI Changes greater than 200 points.\footnote{HHI thresholds are discussed in the 2010 Horizontal Merger Guidelines, Section 5.3.  }

@@

Each panel contains four pairs of box and whisker plots, with each pair corresponding to a different type of merger. The blue box and whisker plots (left) depict outcomes assuming that marginal costs are constant, and the orange box and whisker plots (right) show outcomes assuming that marginal costs are linear.  The whiskers display the $5^{th}$ and $95^{th}$ percentiles of the outcome distribution, the boxes denote the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line marks the median. Note that negative outcome values imply agent harm, and positive values imply agent benefits.

@@

%Consumer harm is less prevalent for mergers where firms have constant marginal costs then when firms have linear costs. Mergers with constant-cost firms benefit consumers in about 15\% of all mergers, while mergers with linear-cost firms benefit consumers in only 2.3\% of mergers. Indeed, consumer harm under linear marginal costs first-order stochastically dominates consumer harm under constant marginal costs. The median consumer harm from markets with linear marginal costs is about 9.3\% of pre-merger total expenditures, about 1.3 times the magnitude of markets with constant marginal costs.

%Consumer harm is less prevalent for mergers where firms have constant marginal costs then when firms have linear costs. Vertical mergers when firms have constant marginal costs benefit consumers in about 30\% of all simulations , but only in about 2.3\% of simulations when firms have linear costs. Upstream mergers with constant marginal cost firms benefit consumers in about 12\% of simulations, but only 0.2\% of simulations when firms have linear costs. Downstream mergers with constant marginal cost firms benefit consumers in about 11\% of simulations, but only about 6\% of simulations when firms have linear costs. Finally, integrated  mergers benefit consumers in about 10\% of all simulations when firms have constant costs, but less than 1\% of simulations when firms have linear costs.

%mergers are unlikely to benefit consumers: upstream under both constant and linear marginal costs, vertical mergers with linear marginal costs, and integrated mergers with linear marginal costs.  That upstream horizontal mergers are largely net harmful is somewhat surprising, as it indicates that any EDM from the unintegrated upstream firm merging with its integrated rival is outweighed by the recapture effect. Similar logic applies to integrated mergers. Likewise, the fact that vertical mergers with linear marginal costs are almost never beneficial when marginal costs are linear is surprising, as it suggests that the incentive to raise rivals' cost dominates the benefit from EDM. %especially given that about 40\% of vertical mergers with constant marginal costs are beneficial.
\textbf{Is it worth exploring which effect dominates: the lessening of EDM or the strengthening of RRC? Maybe do this by allowing assymetric cost structures between merging and non-merging parties?}

@@

Turning to retailers in the second panel of Figure \ref{fig:surplussum}, we find that while downstream mergers, vertical mergers and integrated mergers always benefit retailers, upstream mergers  harm retailers in about 70\% of all simulations. Moreover, there is a partial rank-ordering across mergers: the retailer surplus distribution from integrated mergers first-order stochastically dominates the retailer surplus distribution from downstream mergers, which dominates the retailer surplus distribution from upstream mergers, but not the retailer surplus distribution from vertical mergers.

@@

In terms of total welfare, with the exception of the approximately 10\% of vertical mergers and 8.5\% of integrated mergers with constant costs that are beneficial, our simulated mergers are almost always net harmful. Moreover there is a complete rank ordering of mergers, with total harm from integrated mergers first-order stochastically dominating total harm from downstream mergers, which dominates total harm from upstream mergers, which dominates totals harm from vertical mergers.

@@

Figure \ref{fig:CVvertincumbBW_updownBW} depicts box and whisker plots summarizing consumer harm (blue,left) and total harm (orange, right) as the number of incumbent integrated firms increases from 0 to 6 firms. The box and whisker  plots when the number of  incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2020}\footnote{Important differences between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2020} is that these results exclude downstream markets where prices are set according to a 2nd score auction, but include markets where costs are linear as well as constant.} . For downstream and upstream  mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, and all 3rd parties are unintegrated. By contrast, for vertical mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when a single 3rd party is integrated. Likewise, for integrated mergers, the plots when the number of  incumbent integrated firms equals 2 depict a merger between two integrated firms, and all rivals are unintegrated.

Starting with vertical mergers, absent incumbent integration (0 integrated firms), about 22\% of vertical mergers benefit consumers and about 14\% of vertical mergers increase total surplus.  The presence of a single 3rd party integrated firm (1 integrated firm) decreases the percentage of mergers that benefit consumers to 14.5\% and the percentage of mergers that increase  total surplus to 9.6\%, but leads median harm largely unchanged. Interestingly, while adding a second integrated incumbent to the market also lowers the percentage of beneficial mergers,adding a third integrated firm increases the percentage of mergers that benefits consumers to about 15\%, and the percentage of mergers that increase total surplus to 6.9\%. Adding a fourth integrated incumbent has a similar effect to adding a third incumbent.

downstream mergers (leftmost column), absent incumbent vertical integration , not only are downstream mergers rarely beneficial, but median consumer harm  from downstream mergers is about 9\% of pre-merger revenues while total harm is about 5\% of pre-merger revenues. A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the range of outcomes while also increasing median harm: consumer welfare is positive in more than 13\% of mergers and total welfare is positive in 3.4\%, but median consumer harm increases to 14\% and median total harm increases to 7.6\% of pre-merger revenues. Increasing the number of integrated 3rd parties to 1 (2 integrated firms) shrinks the inter-quartile range but has little impact on median harm. Adding additional integrated 3rd parties (3 integrated firms) further decreases the range of outcomes while also decreasing median harm. With 3 integrated 3rd parties (4 integrated firms), a downstream merger between an integrated and unintegrated firms benefits consumers in about 20\% of markets and shrinks median consumer harm to about 6\% of pre-merger revenues. By contrast, with 3 integrated 3rd parties, the inter-quartile range shrinks, and median total harm falls to about 3.7\% of pre-merger revenues.

Like downstream mergers, not only are upstream mergers absent incumbent vertical integration rarely beneficial, but median consumer harm  from upstream mergers is about 6\% of pre-merger revenues  and total harm is about 3\% of pre-merger revenues. A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the proportion of beneficial mergers but does little to change median harm: consumer welfare is positive in less than 8\% of mergers and total welfare is positive in 2.3\% of mergers, but median consumer harm increases to 7\% and median total harm increases to 4.8\% of pre-merger revenues. Increasing the number of integrated 3rd parties to 1 (2 integrated firms) shrinks the inter-quartile range but has little impact on median harm. Adding additional integrated 3rd parties (3,4 integrated firms) further decreases the range of outcomes while leaving median harm largely unchanged.

Finally, absent the presence of rival incumbent integrated firms, mergers between two integrated firms harm consumers in about 96\% of simulated markets and lowers total surplus in more than 92\% of simulated markets. while adding one additional rival incumbent integrated firm (3 integrated firms) has little effect on either the distribution of consumer or total surplus, additional integrated rivals eliminate the most harmful integrated mergers while also increasing median consumer harm by as much as 13.3\% (5 integrated firms) and median total harm by as much as 8.4\%.

@@

\begin{sidewaysfigure}
\centering
\includegraphics[scale=0.9]{../output/CVvertincumb_updownBW.png}
\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (blue,left) and total (orange,right) surplus as the number of vertically integrated firms present in a market change.  Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median. }
\label{fig:CVvertincumbupdownBW}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\centering
\includegraphics[scale=0.9]{../output/CVvertincumb_vertincBW.png}
\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (blue,left) and total (orange,right) surplus as the number of vertically integrated firms present in a market change.  Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median. }
\label{fig:CVvertincumbvertincBW}
\end{sidewaysfigure}

>>>>

Table \ref{tab:simsum} provides summary statistics across our various simulations.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to calibrate and simulate the effects of mergers in a range of competitive scenarios, including the ones described here.}  The median average wholesale pre-merger price is \$4.6, and the median average retail pre-merger price is \$14.  Because the market size is set to 1, these average prices are equal to total pre-merger expenditures.  Pre-merger HHIs range between 2,814 at the $25^{th}$ percentile to 5,030 at the $75^{th}$, with a median of 3,703. HHIs for vertical mergers increase by 1,277 points at the median, resulting in a median post-merger HHI equal to 4,889.\footnote{We compute the post-merger HHI for vertical mergers by calculating the merged firms' market share as the sum of all the shares of downstream products that either incorporate the upstream partner's input or are sold by the downstream partner.  }  HHIs for upstream mergers increase by 2,004 points at the median, resulting in a median post-merger HHI equal to 5,231. HHIs for horizontal downstream mergers increase by 2,054 points at the median, resulting in a median post-merger HHI equal to 4,889.   HHIs for integrated mergers increase by 2,746 points at the median, resulting in a median post-merger HHI equal to 7,220. Many of these markets fall into the span designated by the DOJ/FTC Horizontal Merger Guidelines as ``Highly Concentrated Markets,'' with post-merger HHIs over 2,500 points and HHI Changes greater than 200 points.\footnote{HHI thresholds are discussed in the 2010 Horizontal Merger Guidelines, Section 5.3.  }

@@

Each panel contains four box and whisker plots, with each plot corresponding to a different type of merger. The whiskers display the $5^{th}$ and $95^{th}$ percentiles of the outcome distribution, the boxes denote the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line marks the median. Note that negative outcome values imply agent harm, and positive values imply agent benefits.

@@

\textbf{Is it worth exploring which effect dominates: the lessening of EDM or the strengthening of RRC? Maybe do this by allowing assymetric cost structures between merging and non-merging parties?}

@@

Turning to retailers in the second panel of Figure \ref{fig:surplussum}, we find that while downstream mergers, vertical mergers and integrated mergers always benefit retailers, upstream mergers  harm retailers in about 70\% of all simulations. Moreover, there is a partial rank-ordering across mergers that is distinct from the consumer rank-ordering: the retailer surplus distribution from integrated mergers first-order stochastically dominates the retailer surplus distribution from downstream mergers, which dominates the retailer surplus distribution from upstream mergers, but not the retailer surplus distribution from vertical mergers.

@@

In terms of total welfare, with the exception of the approximately 10\% of vertical mergers and 8.5\% of integrated mergers with constant costs that are beneficial, our simulated mergers are almost always net harmful. Moreover, there is a partial rank ordering of mergers, with total harm from downstream mergers first-order stochastically dominating total harm from upstream mergers, which dominates totals harm from vertical mergers. Median consumer harm from integrated mergers is about 10\% of pre-merger total expenditures, X.XX times the magnitude of that from downstream mergers, X.X times the magnitude of that from upstream mergers, and more than X.X times the magnitude of vertical mergers.

@@

Figure \ref{fig:CVvertincumbBW_updownBW} depicts box and whisker plots summarizing consumer harm (top) and total harm (bottom) as the number of incumbent integrated firms increases from 0 to 6 firms. The box and whisker  plots when the number of  incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2020}.\footnote{Important differences between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2020} is that these results exclude downstream markets where prices are set according to a 2nd score auction}  For vertical mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when a single 3rd party is integrated. By contrast,for downstream and upstream  mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, with no integrated third parties. Finally, for integrated mergers, the plots when the number of  incumbent integrated firms equals 2 depict a merger between two integrated firms, again with no integrated rivals.

Starting with vertical mergers, absent incumbent integration (0 integrated firms), about 22\% of vertical mergers benefit consumers and about 14\% of vertical mergers increase total surplus.  The presence of a single 3rd party integrated firm (1 integrated firm) decreases the percentage of mergers that benefit consumers to 14.5\% and the percentage of mergers that increase  total surplus to 9.6\%, but leads median harm largely unchanged. Interestingly, while adding a second integrated incumbent to the market also lowers the percentage of beneficial mergers,adding a third integrated firm increases the percentage of mergers that benefits consumers to about 15\%, and the percentage of mergers that increase total surplus to 6.9\%. Adding a fourth integrated incumbent has a similar effect to adding a third incumbent. Taken together, Figure \ref{fig:CVvertincumbBW_updownBW} indicates that the presence of incumbent integrated firms is harmful, and that the harm does not lessen as additional additional integrated incumbents are added.

By contrast, not only are upstream mergers absent incumbent vertical integration rarely beneficial, but median consumer harm  from upstream mergers is about 6\% of pre-merger revenues  and total harm is about 3\% of pre-merger revenues. A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the proportion of beneficial mergers but does little to change median harm: consumer welfare is positive in less than 8\% of mergers and total welfare is positive in 2.3\% of mergers, but median consumer harm increases to 7\% and median total harm increases to 4.8\% of pre-merger revenues. Increasing the number of integrated 3rd parties to 1 (2 integrated firms) shrinks the inter-quartile range but has little impact on median harm. Adding additional integrated 3rd parties (3,4 integrated firms) further decreases the range of outcomes while leaving median harm largely unchanged. In summary, additional incumbent integrated firms do not substantially impact median harm, but do substantially reduce the likelihood of observing extremely harmful mergers.

Like upstream mergers, downstream mergers absent incumbent vertical integration are rarely beneficial, with median consumer harm from downstream mergers equal to about 9\% of pre-merger revenues while total harm is about 5\% of pre-merger revenues. A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the range of outcomes while also increasing median harm: consumer welfare is positive in more than 13\% of mergers and total welfare is positive in 3.4\%, but median consumer harm increases to 14\% and median total harm increases to 7.6\% of pre-merger revenues. Increasing the number of integrated 3rd parties to 1 (2 integrated firms) shrinks the inter-quartile range but has little impact on median harm. Adding additional integrated 3rd parties (3 integrated firms) further decreases the range of outcomes while also decreasing median harm. With 3 integrated 3rd parties (4 integrated firms), a downstream merger between an integrated and unintegrated firms benefits consumers in about 20\% of markets and shrinks median consumer harm to about 6\% of pre-merger revenues. By contrast, with 3 integrated 3rd parties, the inter-quartile range shrinks, and median total harm falls to about 3.7\% of pre-merger revenues.

Finally, absent the presence of rival incumbent integrated firms, mergers between two integrated firms harm consumers in about 96\% of simulated markets and lowers total surplus in more than 92\% of simulated markets. While adding one additional rival incumbent integrated firm (3 integrated firms) has little effect on either the distribution of consumer or total surplus, additional integrated rivals eliminate the most harmful integrated mergers while also increasing median consumer harm by as much as 13.3\% (5 integrated firms) and median total harm by as much as 8.4\%.

@@

% \begin{sidewaysfigure}
% \centering
% \includegraphics[scale=0.9]{../output/CVvertincumb_updownBW.png}
% \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (blue,left) and total (orange,right) surplus as the number of vertically integrated firms present in a market change.  Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median. }
% \label{fig:CVvertincumbupdownBW}
% \end{sidewaysfigure}
%
% \begin{sidewaysfigure}
% \centering
% \includegraphics[scale=0.9]{../output/CVvertincumb_vertincBW.png}
% \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (blue,left) and total (orange,right) surplus as the number of vertically integrated firms present in a market change.  Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median. }
% \label{fig:CVvertincumbvertincBW}
% \end{sidewaysfigure}
